### 2017-11-14 submission/manuscript.Rmd
add Will's suggestions to the results

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. Next, we assessed how typical CRC associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Third, using Random Forest models, we analyzed whether the full community or only the CRC associated genera resulted in better model classification based on the area under the curve (AUC). Since the changes in community were subtle for adenoma we also examined what effect size and sample size the studies that were used were adequately powered for.

@@

***Diversity is Lower in Those with Carcinoma and Increases Relative Risk:***
Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces and not tissue were  lower in those with carcinoma [Figure 1]. Using linear mixed-effect models to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`). This effect was not observed in tissue when additionally controlling for whether the sample came from the same individual (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). For fecal samples a decrease in Shannon diversity and evenness resulted in a significantly increased relative risk for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2]. Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2]. Only a decrease in evenness had an increased relative risk for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))`).

Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls but not between adenoma and controls [Table S1 & S2]. Many studies with unmatched tissue samples had a significant difference for both adenoma and carcinoma versus controls [Table S3 & S4] while studies with matched tissue samples had no differences [Table S3 & S4].

***Genera Previously Associated with Carcinoma Predominately Increases Relative Risk for Carcinoma but not Adenoma:***
The majority of CRC associated genera for both feces and tissue had a significantly increased relative risk for carcinoma but not for adenoma [Figure 3]. In fecal samples the relative risk due to CRC associated genera was greater than either evenness or Shannon diversity [Figure 2 & 3]. Additionally, the relative risk of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D]. The relative risk effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that the tissue analysis include matched samples.

Two measures in stool for adenoma were significant when investigating these CRC associated genera. The first was *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and the second was being positive for three CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A]. For tissue three measures for adenoma were significant; being positive for one CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), being positive for two CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].

***Using the Whole Community Increases Model AUC over CRC Associated Genera:***
For both fecal and tissue samples (matched and unmatched) the AUC decreases when only OTUs from the CRC associated genera are used [Figure 4 & 5]. This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5]. The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S2-S3]. Both genus models perform similarily in their ability to be able to predict lesion (adenoma or carcinoma) with carcinoma having a higher AUC then adenoma [Figure S4-S5]. Matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S4A, S5B, & S6] then carcinoma models [Figure S4B & S5A].

***Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B]. No single study that was analyzed had the standard 80% power to detect an effect size difference that was eqaul to or below 10% [Figure 6A & B].  In order to achieve adequate power for small effect sizes it would be necessary to recruit over 1000 individuals for each arm of the study [Figure 6C].

>>>>

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totalling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether the diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. Next, we assessed how typical CRC associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the relative risk of adenoma or carcinoma. Third, using Random Forest models, we analyzed whether the full community or only the CRC associated genera resulted in better model classification based on the area under the curve (AUC). Since the changes in community were subtle for adenoma we also examined what effect size and sample size the studies that were used were adequately powered for.

@@

***Lower Community Diversity is Associated with Increased RR of Carcinomas:***
Using power transformed and Z-score normalized $\alpha$-diversity metrics, both evenness and Shannon diversity in feces, but not tissue, were  lower in those with carcinoma [Figure 1]. Using linear mixed-effect models to control for study and variable region, there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`). However, in tissue, this effect was not observed in tissue when resampling of the same individual was also controlled for (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). Within fecal samples, a decrease in Shannon diversity and evenness resulted in a significantly increased RR for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2]. Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2]. Only a decrease in evenness had an increased RR for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))`).

Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls, but not adenoma and controls [Table S1 & S2]. For studies with unmatched tissue samples a similar trend was observed [Table S3 & S4] while studies with matched tissue samples had no differences [Table S3 & S4].

***Carcinoma-Associated Genera Minimally Impacts RR of Adenoma:***
The majority of CRC-associated genera for both feces and tissue had a significantly increased RR for carcinoma but not for adenoma [Figure 3]. In fecal samples the RR due to CRC associated genera was greater than either the RR assoicated with evenness or Shannon diversity [Figure 2 & 3]. Additionally, the RR of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) than for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that the tissue analysis included matched samples.

There were two significant measures for increased RR of adenoma when investigating CRC-associated genera in stool: 1) Having a higher then median value of *Porphyromonas* (P-value = `r format(filter(adn_select_RR_stool, measure == "Porphyromonas")[, "pvalue"], digits = 2, nsmall = 2)`) and 2) whether samples were positive for three CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_stool, measure == "three_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3A]. With tissue, there were three significant measures for an increased RR of adenoma: 1) being positive for one CRC-associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "one_counts")[, "pvalue"], digits = 2, nsmall = 2)`), 2) being positive for two CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "two_counts")[, "pvalue"], digits = 2, nsmall = 2)`), and 3) being positive for four CRC associated genera (P-value = `r format(filter(adn_select_inc_4_RR_tissue, measure == "four_counts")[, "pvalue"], digits = 2, nsmall = 2)`) [Figure 3C].

***Using the Whole Community Instead of Only CRC-Associated Genera Increases Model AUC:***
For both fecal and tissue (matched and unmatched) samples, the AUC decreases when only OTUs from the CRC-associated genera are used [Figure 4 & 5]. This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5]. The genus models generally performed similar to the OTU-based models with the full genera models performing better than the CRC-associated genera models [Figure S2-S3]. Both genus models were similarily predictive in their ability to detect adenomas or carcinomas, with carcinoma having a higher AUC then adenoma [Figure S4-S5]. Of note, matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S4A, S5B, & S6] than carcinoma models [Figure S4B & S5A].

***A Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B]. No single study that was analyzed had the standard 80% power to detect an effect size difference that was equal to or below 10% [Figure 6A & B].  In order to achieve adequate power for small effect sizes, studies would need to recruit over 1000 individuals for each arm [Figure 6C].

### 2017-11-14 submission/manuscript.tex
add Will's suggestions to the results

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totalling over 2229 total samples across 14 studies
{[}12--25{]}, we expand both the breadth and scope of the previous
meta-analysis to investigate whether the bacterial community is an
important risk factor for both adenoma and carcinoma. To accomplish this
we first assessed whether the diversity changes throughout disease
(control to adenoma to carcinoma) and if it results in an increased
relative risk (RR) for adenoma or carcinoma. Next, we assessed how
typical CRC associated genera (\emph{Fusobacterium}, \emph{Parvimonas},
\emph{Peptostreptococcus}, or \emph{Porphyromonas}) affect the relative
risk of adenoma or carcinoma. Third, using Random Forest models, we
analyzed whether the full community or only the CRC associated genera
resulted in better model classification based on the area under the
curve (AUC). Since the changes in community were subtle for adenoma we
also examined what effect size and sample size the studies that were
used were adequately powered for.

@@

\textbf{\emph{Diversity is Lower in Those with Carcinoma and Increases
Relative Risk:}} Using power transformed and Z-score normalized
\(\alpha\)-diversity metrics, both evenness and Shannon diversity in
feces and not tissue were lower in those with carcinoma {[}Figure 1{]}.
Using linear mixed-effect models to control for study and variable
region there was a significant decrease from control to adenoma to
carcinoma for both evenness (P-value = 0.025) and Shannon diversity
(P-value = 0.043). This effect was not observed in tissue when
additionally controlling for whether the sample came from the same
individual (P-value \textgreater{} 0.05). For fecal samples a decrease
in Shannon diversity and evenness resulted in a significantly increased
relative risk for carcinoma (P-value = 0.01 and P-value = 0.0011,
respectively) {[}Figure 2{]}. Although these values were significant,
the effect size was relatively small for both metrics (Shannon RR = 1.31
and evenness RR = 1.34) {[}Figure 2{]}. Only a decrease in evenness had
an increased relative risk for adenoma (P-value = 0.032) {[}Figure 2A \&
S1{]} but this effect size was even smaller than what was observed for
carcinoma (RR = 1.16).

Using the Bray-Curtis distance metric, there was a significant
difference across studies in the bacterial community of fecal samples
between carcinoma and controls but not between adenoma and controls
{[}Table S1 \& S2{]}. Many studies with unmatched tissue samples had a
significant difference for both adenoma and carcinoma versus controls
{[}Table S3 \& S4{]} while studies with matched tissue samples had no
differences {[}Table S3 \& S4{]}.

\textbf{\emph{Genera Previously Associated with Carcinoma Predominately
Increases Relative Risk for Carcinoma but not Adenoma:}} The majority of
CRC associated genera for both feces and tissue had a significantly
increased relative risk for carcinoma but not for adenoma {[}Figure
3{]}. In fecal samples the relative risk due to CRC associated genera
was greater than either evenness or Shannon diversity {[}Figure 2 \&
3{]}. Additionally, the relative risk of carcinoma continuously
increased as individuals tested positive for more CRC associated genera
{[}Figure 3B \& 3D{]}. The relative risk effect size was greater for
stool (RR range = 1.61 - 2.74) then for tissue (RR range = 1.21 - 1.81).
This decrease may be explained by the fact that the tissue analysis
include matched samples.

Two measures in stool for adenoma were significant when investigating
these CRC associated genera. The first was \emph{Porphyromonas} (P-value
= 0.023) and the second was being positive for three CRC associated
genera (P-value = 0.022) {[}Figure 3A{]}. For tissue three measures for
adenoma were significant; being positive for one CRC associated genera
(P-value = 0.032), being positive for two CRC associated genera (P-value
= 0.008), and being positive for four CRC associated genera (P-value =
0.039) {[}Figure 3C{]}.

\textbf{\emph{Using the Whole Community Increases Model AUC over CRC
Associated Genera:}} For both fecal and tissue samples (matched and
unmatched) the AUC decreases when only OTUs from the CRC associated
genera are used {[}Figure 4 \& 5{]}. This decrease is observed in both
adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The genus models
generally had similar trends as observed for the OTU based models with
the full genera models performing better then the CRC associated genera
models {[}Figure S2-S3{]}. Both genus models perform similarily in their
ability to be able to predict lesion (adenoma or carcinoma) with
carcinoma having a higher AUC then adenoma {[}Figure S4-S5{]}. Matched
tissue samples for those with carcinoma had an AUC that was more similar
to the adenoma models {[}Figure S4A, S5B, \& S6{]} then carcinoma models
{[}Figure S4B \& S5A{]}.

\textbf{\emph{Majority of Studies are Underpowered for Detecting Small
Effect Size Differences:}} When assessing the power of each study at
different effect sizes the majority of studies for both adenoma and
carcinoma have an 80\% power to detect a 30\% difference {[}Figure 6A \&
B{]}. No single study that was analyzed had the standard 80\% power to
detect an effect size difference that was eqaul to or below 10\%
{[}Figure 6A \& B{]}. In order to achieve adequate power for small
effect sizes it would be necessary to recruit over 1000 individuals for
each arm of the study {[}Figure 6C{]}.

>>>>

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totalling over 2229 total samples across 14 studies
{[}12--25{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope of
the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. To
accomplish this we first assessed whether the diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. Next, we
assessed how typical CRC associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
affect the relative risk of adenoma or carcinoma. Third, using Random
Forest models, we analyzed whether the full community or only the CRC
associated genera resulted in better model classification based on the
area under the curve (AUC). Since the changes in community were subtle
for adenoma we also examined what effect size and sample size the
studies that were used were adequately powered for.

@@

\textbf{\emph{Lower Community Diversity is Associated with Increased RR
of Carcinomas:}} Using power transformed and Z-score normalized
\(\alpha\)-diversity metrics, both evenness and Shannon diversity in
feces, but not tissue, were lower in those with carcinoma {[}Figure
1{]}. Using linear mixed-effect models to control for study and variable
region, there was a significant decrease from control to adenoma to
carcinoma for both evenness (P-value = 0.025) and Shannon diversity
(P-value = 0.043). However, in tissue, this effect was not observed in
tissue when resampling of the same individual was also controlled for
(P-value \textgreater{} 0.05). Within fecal samples, a decrease in
Shannon diversity and evenness resulted in a significantly increased RR
for carcinoma (P-value = 0.01 and 0.0011, respectively) {[}Figure 2{]}.
Although these values were significant, the effect size was relatively
small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
{[}Figure 2{]}. Only a decrease in evenness had an increased RR for
adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect size was
even smaller than what was observed for carcinoma (RR = 1.16).

Using the Bray-Curtis distance metric, there was a significant
difference across studies in the bacterial community of fecal samples
between carcinoma and controls, but not adenoma and controls {[}Table S1
\& S2{]}. For studies with unmatched tissue samples a similar trend was
observed {[}Table S3 \& S4{]} while studies with matched tissue samples
had no differences {[}Table S3 \& S4{]}.

\textbf{\emph{Carcinoma-Associated Genera Minimally Impacts RR of
Adenoma:}} The majority of CRC-associated genera for both feces and
tissue had a significantly increased RR for carcinoma but not for
adenoma {[}Figure 3{]}. In fecal samples the RR due to CRC associated
genera was greater than either the RR assoicated with evenness or
Shannon diversity {[}Figure 2 \& 3{]}. Additionally, the RR of carcinoma
continuously increased as individuals tested positive for more CRC
associated genera {[}Figure 3B \& 3D{]}. The RR effect size was greater
for stool (RR range = 1.61 - 2.74) than for tissue (RR range = 1.21 -
1.81). This decrease may be explained by the fact that the tissue
analysis included matched samples.

There were two significant measures for increased RR of adenoma when
investigating CRC-associated genera in stool: 1) Having a higher then
median value of \emph{Porphyromonas} (P-value = 0.023) and 2) whether
samples were positive for three CRC associated genera (P-value = 0.022)
{[}Figure 3A{]}. With tissue, there were three significant measures for
an increased RR of adenoma: 1) being positive for one CRC-associated
genera (P-value = 0.032), 2) being positive for two CRC associated
genera (P-value = 0.008), and 3) being positive for four CRC associated
genera (P-value = 0.039) {[}Figure 3C{]}.

\textbf{\emph{Using the Whole Community Instead of Only CRC-Associated
Genera Increases Model AUC:}} For both fecal and tissue (matched and
unmatched) samples, the AUC decreases when only OTUs from the
CRC-associated genera are used {[}Figure 4 \& 5{]}. This decrease is
observed in both adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The
genus models generally performed similar to the OTU-based models with
the full genera models performing better than the CRC-associated genera
models {[}Figure S2-S3{]}. Both genus models were similarily predictive
in their ability to detect adenomas or carcinomas, with carcinoma having
a higher AUC then adenoma {[}Figure S4-S5{]}. Of note, matched tissue
samples for those with carcinoma had an AUC that was more similar to the
adenoma models {[}Figure S4A, S5B, \& S6{]} than carcinoma models
{[}Figure S4B \& S5A{]}.

\textbf{\emph{A Majority of Studies are Underpowered for Detecting Small
Effect Size Differences:}} When assessing the power of each study at
different effect sizes the majority of studies for both adenoma and
carcinoma have an 80\% power to detect a 30\% difference {[}Figure 6A \&
B{]}. No single study that was analyzed had the standard 80\% power to
detect an effect size difference that was equal to or below 10\%
{[}Figure 6A \& B{]}. In order to achieve adequate power for small
effect sizes, studies would need to recruit over 1000 individuals for
each arm {[}Figure 6C{]}.
